Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 paper.tex
updating intro

However, MHD turbulence may not be applicable for all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, other mechanisms 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. This
height-dependence, or vertical shear, enables perturbations with
large vertical lengthscales (relative to radial lengthscales) to
become unstable. The VSI is the disk analog of the
Goldreich-Schubert-Fricke instability
\citep[GSF,]{goldreich67,fricke68} in differentially rotating stars.   
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

The VSI has

@@

recently been demonstrated through 
hydrodynamical simulations \citep{nelson13}

%and kley simulations

%vsi background

%questions for this work

@@

>>>>

However, MHD turbulence may not be applicable for all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, alternative routes 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}, where a Fourier analysis was performed in 
both the radial and vertical directions. Simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13}. In these simulations the authors indeed observe the
growth of vertically-global but radially-localized perturbations when
the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P$, \cite{nelson13} find that
$t_c\lesssim 0.01P$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the 
VSI is only sustained in radiation hydrodynamic simulations 
with the inclusion of heating from stellar irradiation.

@@

From these works it is clear that the VSI is sensitive 
to the thermodynamic response of the disk except in the special case of neutrally
stratified disks. It is therefore important to understand this dependence in order to evaluate
the viability of the VSI in realistic disks. However,
